Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-13 aistats.tex
Start rephrasing the open problems

Exponential families are among the most widely used statistical models.
Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
They are both flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
In particular, they are appreciated for their convex log-likelihood
\alignn{
f(\nat) = \E[-\log p(X \cond \nat)] = \logpart(\nat) - \lin{\E[T(X)] , \nat}
}
where $\logpart$ is a convex function and $\nat$ is a parameter vector.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} 
or variants of principal component analysis \citep{collins2001generalization} among other applications.
Another interest of exponential families is their closed form maximum-likelihood estimate (MLE) obtained via moment matching
\begin{align}
	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} 
	\label{eq:defMLE}
\end{align}
Given a conjugate prior, a similar formula holds for the maximum a posteriori (MAP).

@@

This question has connections with optimization.
For one, the MLE is minimizing $f$ (or a regularized version for MAP), and the KL is the sub-optimality
\alignn{
f(\nat) - f(\nat^*) = \KL\paren{p(\cdot \cond \nat^*) || p(\cdot \cond \nat) }
}
where $\nat^* \in \argmin f(\nat)$.
For two, in exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
A convergence rate for this algorithm would translate into the upper bound we seek.
%similarly to other works on stochastic approximation \citep{bach2013nonstronglyconvex}. %TODO check this
From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g. not well approximated by quadratics), but benefits from \emph{relative} smoothness properties \citep{birnbaum2011distributed}.
In fact we can see this problem as a very special case to study SMD, revealing the limits of current analysis.

@@

The exponential family for data $X\in \cX$ reads
\begin{equation}
	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; ,
	 \label{eq:def_expfamily}
\end{equation}
where  $\nat$ is called natural (or primal) parameter is specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) $\nu(dx)$, the density with respect to a base measure.
Since the exponential is positive, $p$ has the same support as $\nu$.
The function $\logpart$ is called the log-partition function that acts as a normalization term, since
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
\end{align}

@@

For convenience, in this paper we focus on steep, regular exponential families
where $\logpart$ is strictly convex function of Legendre type,
and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
(See \citet{barndoffnielsen2014information}). Also, when explicit, we write the random variable $T = T(X)$.

@@

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p(T|\nat)]{T} =: \meanp \\
    \nabla^2 \logpart(\nat) &= \Cov_\nat[T] \geq 0
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both represent the same distribution.

We now introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\m) = \min_{\nat\in\Theta}\langle \m, \nat \rangle - \logpart(\nat) \; ,
\end{align}
which matches the common notion of \textit{entropy} in information theory. In this paper, we assume that the sufficient statistic $T$ is minimal, e.g., its covariance is definite positive for all $\nat$. In such case, the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$, and moreover, $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}). More concisely,
\[
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
\]

% The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).

\begin{figure}[t] % @remi In the figure you use mu instead of m??
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
	}
	\label{fig:duality}
\end{figure}

\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps)
We now introduce the Bregman divergence induced by $\logpart$, that measure the distance between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. In general, bregman distances are non-symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$. Finally, it is equal to the divergence of its convex conjugate with switched arguments,
\begin{align}
	\bregman (\nat ; \nat_0)
    = \bregmanconj ( \meanp_0 ; \meanp) \; .
\end{align}

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
    \label{eq:def_prior}
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is the number of fictive data points observed from a distribution with parameter mean $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).} 
\textcolor{red}{\textbf{D.S.: This section is very unclear. }}
Using Bayes rule with~\eqref{eq:def_expfamily}~\&~\eqref{eq:def_prior}, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat \cond \mD_n)
	%\propto p(\mD|\nat)p(\nat)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE).
In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
The MLE and MAP estimates are statistics of the dataset $\mD_n$.
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

\section{Open Problem}
Given a distribution $\cD$, our objective is the expected log-likelihood 
\begin{align}
	f(\nat) = \expect[X \sim \cD]{-\log p(X \cond \nat)} 
	 = \logpart(\nat) - \langle \E[T] , \nat \rangle
\end{align}
Note that $\cD$ does not have to be part of the exponential family.
Let $\nat^* := \argmin f(\nat)$.
Assuming $\nat^*$ exists within $\Theta$ means $\cD$ is such that $\E[T] = \meanp^* \in \cM$.
Then the suboptimality on the population log-likelihood can be written with Bregman divergences
\footnote{
Since $f$ is also the KL (up to a constant)  $f(\nat) + \cst = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$, 
and for exponential families $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, we recover the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
}
\begin{align}
	 f(\nat) - f(\nat^*)
	 %&= \logpart(\nat) - \logpart(\nat^*) - \langle \E[T] , \nat - \nat^* \rangle \\
	 = \bregman(\nat ; \nat^*) 
	 = \bregmanconj(\m^* ; \m) \; .
\end{align}
Our question is: how does this suboptimality behave when $\nat$ is the MLE or the MAP ? Can we get upper-bounds in expectations ? e.g. on these quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
More explicitly, can we write an upper bound that does not involve this expectation over the dataset.

@@

{\bf Remark.}
This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
However there is a connection between the random variable $T(X)$ and the metric $\logpart$.
Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable.
For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.

{\bf Remark 2.}
The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$. Instead of taking the expectation,  we might want to bound this quantity in high probability, without resorting to Markov inequality, but that is a difficult endeavor.

>>>>

Exponential families are among the most widely used statistical models.
Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
They are both flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
In particular, they are appreciated for their convex log-likelihood
\alignn{
f(\nat) = \E[-\log p(X \cond \nat)] = \logpart(\nat) - \lin{\E[T(X)] , \nat}
\label{eq:defNLL}
}
where $\logpart$ is a convex function and $\nat$ is a parameter vector.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} 
or variants of principal component analysis \citep{collins2001generalization} among other applications.
Another interest of exponential families is their closed form maximum-likelihood estimate (MLE) obtained via moment matching
\begin{align}
	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} 
	\label{eq:defMLE}
\end{align}
Given a conjugate prior, a similar formula holds for the maximum a posteriori (MAP).

@@

This question has connections with optimization.
For one, the MLE is minimizing $f$ (or a regularized version for MAP), and the KL is the sub-optimality
\alignn{
f(\nat) - f(\nat^*) = \KL\paren{p(\cdot \cond \nat^*) || p(\cdot \cond \nat) }
\label{eq:suboptimalityKL}
}
where $\nat^* \in \argmin f(\nat)$.
For two, in exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
A convergence rate for this algorithm would translate into the upper bound we seek.
%similarly to other works on stochastic approximation \citep{bach2013nonstronglyconvex}. %TODO check this
From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g. not well approximated by quadratics), but benefits from \emph{relative} smoothness properties \citep{birnbaum2011distributed}.
In fact we can see this problem as a very special case to study SMD, revealing the limits of current analysis.

@@

The density of an exponential family for data $x \in \cX$ reads
\begin{equation}
	 p_\nat(x) = p(x|\nat) = \exp( \langle \nat, T(x) \rangle - \logpart(\nat)) \; ,
	 \label{eq:def_expfamily}
\end{equation}
where  $\nat$ is called natural (or primal) parameter.
It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) a base measure$\nu$ on $\cX$.
Since the exponential is positive, $p$ has the same support as $\nu$.
The function $\logpart$ is called the log-partition function that acts as a normalization term, since
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
\end{align}

@@

For convenience, in this paper we focus on steep, regular exponential families with minimal statistic $T$.
Then $\logpart$ is a strictly convex function of Legendre type,
and the set $\Theta = \{ \nat \cond \logpart(\nat) < \infty\}$ is open and convex
\citep{barndoffnielsen2014information}. 
Also, when explicit, we write the random variable $T = T(X)$.

@@

\paragraph{Duality}
The log-partition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
Given that $\logpart$ is strictly convex, its Hessian is positive definite and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both are linked and represent the same distribution.

We now introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the log-partition function
\begin{align}
	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
\end{align}
which matches the common notion of \textit{entropy} in information theory.
By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}). More concisely,
\aligns{
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
}

\begin{figure}[t] 
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
	}
	\label{fig:duality}
\end{figure}

\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
We now introduce the Bregman divergence induced by $\logpart$, that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. In general, bregman distances are non-symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$. Finally, it is equal to the divergence of its convex conjugate with switched arguments,
\begin{align}
	\bregman (\nat ; \nat_0)
    = \bregmanconj ( \meanp_0 ; \meanp) \; .
    \label{eq:bregman_switch}
\end{align}

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
    \label{eq:def_prior}
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is the number of fictive data points observed from a distribution with natural parameter $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).} 
Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD|\nat)p(\nat)$.
Plugging in~\eqref{eq:def_expfamily} and ~\eqref{eq:def_prior} yields
\begin{align}
	p(\nat \cond \mD_n)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
which reaches its maximum in $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)~ \eqref{eq:defMLE}.
In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
%The MLE and MAP estimates are statistics of the dataset $\mD_n$.

\section{Open Problem}
Let us assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
As we will see, this is already an interesting setting.
We wish to estimate how well the MLE or a MAP model the true distribution.

A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$.
In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
Using~\eqref{eq:defNLL} and~\eqref{eq:bregman_switch}, we can also write the KLwith Bregman divergences
\alignn{
	\KL(p_{\nat^*} || p_\nat)
	 = \bregman(\nat ; \nat^*) 
	 = \bregmanconj(\m^* ; \m) \; .
}

In the words of statistical decision theory would refer to these quantities as \emph{frequentist risk} of the estimators, when the statistical loss is given by the divergence the KL.

Given a distribution $\cD$, our objective is the expected log-likelihood 
\begin{align}
	f(\nat) = \expect[X \sim \cD]{-\log p(X \cond \nat)} 
	 = \logpart(\nat) - \langle \E[T] , \nat \rangle
\end{align}
Note that $\cD$ does not have to be part of the exponential family.
If $\cD$ is such that $\E[T] = \meanp^* \in \cM$, then $f(\nat)$ reaches its minimum $\nat^*\in\Theta$.
Then the suboptimality on the population log-likelihood can be written with Bregman divergences
%\footnote{
%Since $f$ is also the KL (up to a constant)  $f(\nat) + \cst = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$, 
%and for exponential families $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, we recover the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
%}
\begin{align}
	 f(\nat) - f(\nat^*)
	 %&= \logpart(\nat) - \logpart(\nat^*) - \langle \E[T] , \nat - \nat^* \rangle \\
	 = \bregman(\nat ; \nat^*) 
	 = \bregmanconj(\m^* ; \m) \; .
\end{align}
Our question is: how does this suboptimality behave when $\nat$ is the MLE or the MAP ? Can we get upper-bounds in expectations ? e.g. on these quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
More explicitly, can we write an upper bound that does not involve this expectation over the dataset.

@@

{\bf Remark.}
While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, for instance in Gaussian covariance estimation when $n \leq d + 1$.
Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality.
